Closing the gap between statistical and scientific workflows for improved forecasts in ecology

### 2025-03-24 forecastflows.tex
add some ideas for current limitations in PBM workflow

\begin{enumerate}
\item Problem
\begin{enumerate}
\item Climate change, biodiversity crisis etc. has made it critical to understand trends to date and be able to forecast future trends (for policy)
\item Current workflows in ecology are not up to the task
\begin{enumerate}
\item For trends: lots of different stuff reported for seemingly same data/question, makes people outside ecology wonder if we can even well document them, let alone understand them enough to suggest policy
\item For forecasting
\end{enumerate}
\item Here we outline the problem and provide a solution!
\end{enumerate}
\item What are current workflows and where are they limiting us?
\begin{enumerate}
\item For trends ...
\begin{enumerate}
\item easy to find different trends through small model tweaks to analyses and/or different data
\item For example, right now many different papers report different biodiversity trends (LPI example?)
\item New workflow would make ecologists understand uncertainty in their model data/combo (and perhaps not see/publish results as so divergent?)
\end{enumerate}
\item For forecasting ... (somehow jump to our focus on process-based models PBMs here? Something like, forecasting is big and there are diverse methods! Near-term iterative, correlative niche models, but PBMs are often considered the gold standard ... mention machine learning?)
\end{enumerate}
\item Better workflows to the rescue! 
\begin{enumerate}
\item General overview of new workflows
\begin{enumerate}
\item Step 0: Research Qs and hypotheses (with a mechanism) lets you ...
\item Step 1: Build a model!
\item Step 2: Simulate data (and priors or something like priors that forces you to put numbers on stuff)
\item Step 3: Design experiments/data collection (maybe you go back to Step 1 here)
\item Step 4: Simulate data from actual design/collection
\item Step 5: Fit the model to empirical data
\item Step 6: Retrodictive checks (feed back to 0 and 1)
\end{enumerate}
\item New vision of each workflow
\end{enumerate}
\item Conclusion: world is better
\end{enumerate}

>>>>

\begin{enumerate}
\item Problem
\begin{enumerate}
\item Climate change, biodiversity crisis etc. has made it critical to understand trends to date and be able to forecast future trends (for policy)
\item Current workflows in ecology are not up to the task
\begin{enumerate}
\item For trends: lots of different stuff reported for seemingly same data/question, makes people outside ecology wonder if we can even well document them, let alone understand them enough to suggest policy
\item For forecasting: also high divergence between models, criticism against process-based approaches (supposedly the most robust?) for lack of transparency (in the model building and calibration) and increasing complexity, with many parameters not supported by data
\end{enumerate}
\item Here we outline the problem and provide a solution!
\end{enumerate}
\item What are current workflows and where are they limiting us?
\begin{enumerate}
\item For trends ...
\begin{enumerate}
\item easy to find different trends through small model tweaks to analyses and/or different data
\item For example, right now many different papers report different biodiversity trends (LPI example?)
\item New workflow would make ecologists understand uncertainty in their model data/combo (and perhaps not see/publish results as so divergent?)
\end{enumerate}
\item For forecasting ... (somehow jump to our focus on process-based models PBMs here? Something like, forecasting is big and there are diverse methods! Near-term iterative, correlative niche models, but PBMs are often considered the gold standard ... mention machine learning?)
\begin{enumerate}
\item as many models as researchers working on process-based models 
\item + accumulation of successive layers in the development of models = significant challenge to scientific transparency, reproducibility, interpretability\\
models often draw inspiration from each other, which is good (way to do science), but not always explicit... (some issues: arbitrarily established parameter value in one model then transmitted to multiple models)
\item focus of researchers: always integrate new mechanisms, new parameters, to increase "realism"...  they intuitively "feel" what kinds of adjustments is needed... but opaque from an external perspective ("black box" of model building and calibration)
\item models rarely fitted as a whole, dozens of parameters without explicitly quantifying parameter uncertainty, and often neglect to propagate this uncertainty
\item simulations of models themselves became a subject of study to disentangle all the processes modelled and understand model sensitivity 
\end{enumerate}
\end{enumerate}
\item Better workflows to the rescue! 
\begin{enumerate}
\item General overview of new workflows
\begin{enumerate}
\item Step 0: Research Qs and hypotheses (with a mechanism) lets you ...
\item Step 1: Build a model!
\item Step 2: Simulate data (and priors or something like priors that forces you to put numbers on stuff)
\item Step 3: Design experiments/data collection (maybe you go back to Step 1 here)
\item Step 4: Simulate data from actual design/collection
\item Step 5: Fit the model to empirical data
\item Step 6: Retrodictive checks (feed back to 0 and 1)
\end{enumerate}
\item New vision of each workflow
\end{enumerate}
\item Conclusion: world is better
\end{enumerate}
